\plaintitle

### 2016-09-20 Paper/proceedings.tex
Comments on related work.

While there is a great deal of foundational work in visualization and graphical perception dealing with the estimation of individual values in visualizations (such as the height of the bar in a bar chart, or the angle of a line in a linegraph), there is comparatively little work on how viewers of visualizations perceive aggregate statistical quantities. Ariely~\cite{ariely2001seeing} suggests that, in concert with the perception of individual objects, we also collect information about the \emph{ensemble} properties of visual displays. Szafir et al.~\cite{szafir2016four} suggest that this \emph{ensemble coding} can afford the accurate estimation of \emph{summary statistics} in visualizations. However, visualizations with good performance for summary tasks may not result in good performance at point tasks, and vice versa~\cite{albers2014task,fuchs2013evaluation}. A further difficulty is that tasks requiring estimations of values in visualizations (where there is a single correct answer) are qualitatively different from tasks requiring predictions (where differing mental models and priors can results in a multitude of potential valid responses).

Scatterplots are a standard design for visualization bivariate data, with a multitude of design parameters that affect their suitability for aggregate tasks~\cite{cleveland1984many}. Prior work has confirmed that viewers can make use of scatterplots to perform prediction tasks (which tacitly rely on estimation of trend) in ways that are robust to both noise~\cite{harvey1997effects} and problem frame~\cite{lewandowsky2011popular}. However, the heuristics used when performing these prediction tasks (such as the anchor-and-adjust method~\cite{bolger1993context}) can introduce biases. Similarly, the design of scatterplots (for instance, the decision to encode class membership with color or with shape) can also impact performance at aggregate tasks~\cite{gleicher2013perception,lewandowsky1989discriminating}. The aspect ratio of graphs can also bias judgments about trend~\cite{beattie2002impact}--- wider aspect ratios can cause viewers to underestimate the severity of effects, when compared to narrower charts. Another bias in prediction tasks is the ``within-the-bar'' bias \cite{newman2012bar}: for visually asymmetric visualizations such as bar charts, points contained within the visual area of the bar glyph are perceived as likelier than those outside the glyph.

Recent work in the visualization community has focused on the perception of correlation in scatterplots. Rensink et al.~\cite{rensink2010perception} show that viewers can estimate correlation with some accuracy in scatterplots. Harrison et al. \cite{harrison2014ranking} extend this finding to other visualization types, although a reanalysis by Kay \& Heer~\cite{kay2016beyond} indicates that, for many of the more esoteric bivariate visualizations, performance at this task is poor. Estimation of correlation can also be biased by the design of scales~\cite{cleveland1982variables}--- the whitespace and aspect ratio changes introduced by expanding the scale of the axes can cause overestimation of correlation.

Our task of trend estimation combines elements of both prediction and correlation estimation tasks. As with prediction, there is not necessarily an unambiguous correct estimation of trend (different modeling and regression methods can produce different trend lines). As with correlation, viewers must make holistic judgments about the dataset in a way that (as per Harrison et al.) likely relies on a set of key visual proxies. And, as with both tasks, we believe that trend estimation can be biased through conscious or unconscious design choices.

>>>>

While there is a great deal of foundational work in visualization and graphical perception dealing with the estimation of individual values in visualizations (such as the height of the bar in a bar chart, or the angle of a line in a line graph), there is comparatively little work on how viewers of visualizations perceive aggregate statistical quantities.

Ariely~\cite{ariely2001seeing} suggests that, in concert with the perception of individual objects, we also collect information about the \emph{ensemble} properties of visual displays. Szafir et al.~\cite{szafir2016four} note that this \emph{ensemble coding} might afford relatively accurate estimation of \emph{summary statistics} in visualizations. However, visualizations with good performance for summary tasks may not result in good performance at point tasks, and vice versa~\cite{albers2014task,fuchs2013evaluation}. A further difficulty is that tasks requiring estimation of values in visualizations (where there is a single correct answer) are qualitatively different from tasks requiring predictions (where differing mental models and priors can result in a multitude of potential valid responses).

Scatter plots are a standard means of visualizing bivariate data, with a multitude of design parameters that affect their suitability for aggregate tasks~\cite{cleveland1984many}. Prior work has confirmed that viewers can make use of scatter plots to perform prediction tasks (which tacitly rely on estimation of trend) in ways that are robust to both noise~\cite{harvey1997effects} and problem frame~\cite{lewandowsky2011popular}. However, the heuristics used when performing these prediction tasks (such as the anchor-and-adjust method~\cite{bolger1993context}) can introduce biases.

Similarly, visual design choices (for instance, the decision to encode class membership with color or with shape) can also impact performance at aggregate tasks~\cite{gleicher2013perception,lewandowsky1989discriminating}. The aspect ratio of graphs can also bias judgments about trends~\cite{beattie2002impact}\,---\,wider aspect ratios can cause viewers to underestimate the severity of effects, when compared to narrower charts. Another bias in prediction tasks is the ``within-the-bar'' bias \cite{newman2012bar}: for visually asymmetric visualizations such as bar charts, points contained within the visual area of the bar glyph are perceived as likelier than those outside the glyph.

Recent work in the visualization community has focused on the perception of correlation in scatter plots. Rensink et al.~\cite{rensink2010perception} show that viewers can estimate correlation with some accuracy in scatter plots. Harrison et al.~\cite{harrison2014ranking} extend this finding to other visualization types, although a re-analysis by Kay \& Heer~\cite{kay2016beyond} indicates that, for many of the more esoteric bivariate visualizations, performance at this task is poor. Estimation of correlation can also be biased by the choice of scales~\cite{cleveland1982variables}\,---\,the whitespace and aspect ratio changes introduced by expanding the scale of the axes can cause overestimation of correlation.

Our task of trend estimation combines elements of both prediction and correlation estimation tasks. As with prediction, there is not necessarily an unambiguous correct estimation of trend (different modeling and regression methods can produce different trend lines). As with correlation, viewers must make holistic judgments about the dataset in a way that (as per Harrison et al.~\cite{harrison2014ranking}) likely relies on a set of key visual proxies. \jeff{What do you mean by ``key visual proxies''? Be clearer so that readers can follow along.} And, as with both tasks, we believe that trend estimation can be biased through conscious or unconscious design choices.
\jeff{Such as? The related work could end on a stronger note if we put forward some concrete hypotheses and note that prior work has not addressed them...}
